Statistical Grand Slams

### 2023-12-12 Lahman Analysis/Formal Analysis/Draft1.qmd
Finished Presentation

Similarly $\eta_h$ is the effect that a player's age squared has on the probability of a HR in an AB. The purpose of this squared term is to capture the non-linearity of HR as player's age. It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a non-linear effect of age on a player's ability to hit HR's. This ability appears to peak about when the player is 30-years-old. We include $\eta_h$ in the model to capture this downward trend that we see after a player reaches 30-years-old. It is nearly impossible to interpret what $\eta_h$ is doing because it is the multiplicative effect of the player's centered age squared. This begs the question, what even is a squared age? How does one interpret it? There is not a clear answer on this question, so it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with what the actual estimates are. However, it is easy to justify our choice for the priors on $\eta_h$, $\mu_2$ and $\sigma_2$. We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero. Further, any distinction to the slightly negative of zero would be hard to justify and, in essence, arbitrary. Therefore, we went with the mean of 0 and the standard deviation of 0.01 to account for any players who have a more or less extreme non-linearity to their trajectories. That is, higher or lower peaks and troughs throughout their career for their probability $\pi$. We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to. Furthermore, the exponential distribution with a parameter of 100 is very concentrated. Since we know that the coefficient of $\eta_h$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100). Thus, we can be confident that our standard deviation will end up being small for our estimate.

>>>>

Similarly $\eta_n$ is the effect that player $n$'s squared age has on his probability of hitting a HR in an AB. The purpose of this squared term is to capture the non-linear relationship between hitting a HR and player $n$'s age. It is apparent in both @fig-HR.By.Age.20to40 and @fig-HR.By.Age.All that there is a parabolic relationship between age and players' ability to hit HR's. This ability appears to peak when the player is roughly 30-years-old. We include $\eta_n$ in the model to capture this peak when a player reaches 30-years-old. It is nearly impossible to interpret what $\eta_n$ is doing because it describes the effect of centered age squared on HR probability $\pi$, and centered age squared on its own is already almost a meaningless concept. Therefore, it is best understood that this term simply accounts for the non-linearity we see in the data and not become too concerned with understanding its estimates. However, it is easy to justify our choice for the priors on $\eta_n$, $\mu_2$ and $\sigma_2$. We chose $\mu_2\sim Normal(0,0.01)$ because we assumed that there would be a negative effect so small that it would be very close to zero. Further, any distinction to the slightly negative of zero would be hard to justify and, in essence, arbitrary. Therefore, we went with the mean of 0 and the standard deviation of 0.01 to account for any players who have a more or less extreme non-linearity to their trajectories. That is, higher or lower peaks and troughs throughout their career for their probability $\pi$. We chose $\sigma_2\sim Exponential(100)$ because standard deviations cannot be negative, which the exponential distribution lends itself nicely to. Furthermore, the exponential distribution with a parameter of 100 is very concentrated. Since we know that the coefficient of $\eta_n$ is likely to be very small since it is being multiplied by the age squared (which can be up over 100). Thus, we can be confident that our standard deviation will end up being small for our estimate.

### 2023-12-12 Lahman Analysis/Formal Analysis/Pres.qmd
Finished Presentation

---
title: "Bayesian Data Analysis"
subtitle: "MATH 470 Summary"
author: "Sam Turner"
format: 
  revealjs:
    theme: white
    embed-resources: true
editor: visual
---

@@

    -   Expect average player's HR probability to be 0.0015

@@

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

@@

HR probability $\pi$ is measured on the logistic scale.

@@

@@

## Model (Again)

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\log(\frac{\pi_{nip}}{1-\pi_{nip}}) &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

-   There are $1.17\times10^{12}$ parameters for this model

>>>>

---
title: "Statistical Grand Slams"
subtitle: "A Bayesian Approach to Modeling Home Run Production in Major League Baseball"
author: "Sam Turner"
format: 
  revealjs:
    theme: white
    embed-resources: true
editor: visual
---

## Introduction

-   In this research project, Dr. Parson and I sought to predict the home run (HR) production of Major League Baseball hitters.

-   If you work for a MLB team, predicting HR's is important because it is the pinnacle outcome of an at-bat (AB)

-   And even if you don't work for a MLB team, you could make a lot of money in Vegas if you can accurately predict HR's!

## Introduction Cont.

-   A problem with predicting player HR's is that there is often limited data available

    -   For example, players who have played for more than 6 seasons are considered very "veteran" but only have 6 player-season data points

-   Generalized linear models (GLM's) from classical statistics have a hard fitting accurate models given limited data points - such as the 6 data point scenario mentioned above

-   This is where a Bayesian model shines - Bayesian models can "learn" from the data improving the effective observations it has for fitting

@@

    -   Expect average player's HR probability to be 0.00015

@@

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\frac{\pi_{nip}}{1-\pi_{nip}} &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

@@

The amount of AB's in a year is given for our prediction of HR's, but we say that the probability that the AB results in a HR, $\pi$ varies according to a number of factors.

## Model Cont.

We can both simulate and predict the amount of HR's of a player. For example, let's examine a player that has 100 AB's and a probability $\pi$ of 0.03.

That is, $HR \sim Normal(100,0.03)$,

Simulation:

```{r}
#| echo: true
set.seed(1)
rbinom(5, 100, 0.03)
```

Prediction:

$E(HR)=AB\cdot \pi=100\cdot0.03=3$

## Data

Our data for the model comes from the Lahman data set. Lahman covers a variety of information on each player, including but not limited to, batting, fielding, team, and player statistics.

In order to be considered in our analysis a player must have accumulated at least 6 seasons of play, with at least 50 AB's in each, and have played from 1973-2019.

@@

## Multilevel Trade-Off - No Pooling

No pooling would mean that each player gets their own posterior $\alpha$ distribution. That is,

$$
\alpha_n\sim Normal(-3.5,1).
$$

This prior would be multiplied by the data to generate the posterior for each player. However, this disadvantages players with less seasons because there is not as much data to go off of. And we know that we could look a player, $m$, who performs similar to $n$ WLOG and generate more accurate predictions.

## Multilevel Trade-Off - Total Pooling

In a total pooling scenario $\alpha$ would just be a dummy variable in the model of the model,

$$
HR=\tau+\Gamma\cdot AB+\alpha_n+\beta_n+\eta_n+\delta_p+\xi_i+\epsilon
$$

But then we run into the problem of multicolinearity. That is, how do we tease out the relationship between $\alpha_n, \delta_p$, and $\xi_i$ since they all are intercept terms? This model also has a hard time allowing for players to stray from the mean for any parameter.

In essence, we lose information from both no and total pooling.

## "Innate" HR Hitting Ability - $\alpha_n$ (Again)

@@

## Model (Again I)

$$
\begin{align}
HR_{nip} &\sim Binomial(AB_{nip},\pi_{nip})\\
\frac{\pi_{nip}}{1-\pi_{nip}} &= \alpha_{n}+\beta_n\cdot (Age_{ni}-30)+\eta_n\cdot (Age_{ni}-30)^2\\
&+\delta_p+\xi_i
\end{align}
$$

-   There are 2,116 parameters for this model

    -   (3×657)+88+47+10

-   This model is classically non-identifiable because of 3 intercept terms!

## Model (Again II)
